Supplemental Materials: \\ \emph{Surface-driven registration method for the structure-informed segmentation of diffusion MR images}

### 2016-02-03 Publications/2015-NeuroImage/responseLetter2.tex

\section*{Response letter to the manuscript no. NIMG-15-2877}

\bigskip
\noindent We thank the editors and the reviewers for revising our manuscript no. NIMG-15-2877 submitted to NeuroImage.
We appreciate the comments made by both reviewers, and considered them fully to improve our paper.
Please, find in the following paragraphs our point-by-point answer to the concerns raised by the reviewers,
  along with the description of the corresponding changes in the manuscript.
Text modifications in the manuscript are highlighted with a new font color, and attached notes indicate 
  which reviewer and comment the corresponding change is addressing to.

\bigskip
\bigskip
\subsection*{Reviewer \#1:}
\begin{reviewpoint}
This paper presents a method for simultaneous registration and segmentation of diffusion images in the presence of geometric distortions. The method uses an anisotropically regularised active contour method to coalign tissue segmentations generated from T1 images to distorted multichannel diffusion data. It has the advantage that it has no requirements for extra scans to be acquired for correction of the distortion. The method demonstrates improved performance over a previous b0--T2 registration approach for the alignment of tissue segmentation maps. Improvements are demonstrated on simulated data, and real data with simulated distortions. The advantage of these results is that the ground truth is known.

I am satisfied that the active contour alignment approach used to align the surfaces is sound, and that the results show theoretic improvements over the T2B method. I am convinced that the method approach is effective for the alignment task. In general the methods are well explained with the exception of a few minor points (below).
\end{reviewpoint}
\reply{%
We thank the reviewer for the time dedicated in his/her review and these positive comments.
We are pleased to confirm that this summary of the paper matches the idea we want to transmit.

@@

\begin{reviewpoint}
My main concerns lie with whether the results in the paper prove what the authors set out in their abstract. Specifically, a stated goal is for the use of anatomically constrained tractography. However, there are no results that show this method would improve connectome extraction. In particular, the authors state that ``this method avoids any requirement for collecting additional scans nor resampling for diffusion unwarping''. I see this has advantages, but what about the consequences of not unwarping on the diffusion tracts themselves see [1][2][3]. In my opinion the authors should address this point possibly with additional results to show how they intend this to be used within a tracking framework.

[1] Jones, Derek K., and Mara Cercignani. ``Twentyfive pitfalls in the analysis of diffusion MRI data.'' NMR in Biomedicine 23.7 (2010): 803820.

[2] Lee, J., et al. ``Correction of Bo EPI distortions in diffusion tensor imaging and white matter tractography.'' Proc. Int. Soc. Magn. Reson. Med. Vol. 11. 2004.

[3] Andersson, J. L., et al. ``Effects of susceptibility distortions on tractography.''Proceedings of International 
Society of Magnetic Resonance in Medicine. Vol. 11. 2004.
\end{reviewpoint}
\reply{%
We apologize since the paper could be misleading in its former version, driving
  the reader to identify its goal restricted to support anatomically constrained
  tractography.
We have modified the last sentence in the abstract and lines 57--58 (introduction) to
  clarify that this method is intended to support any diffusion processing task that
  requires a precise delineation of the anatomy in the diffusion space.
In particular, although we have not found the exact phrase cited by the
 reviewer\footnote{``this method avoids any
 requirement for collecting additional scans nor resampling for diffusion unwarping''},
 we have revised our close statements in lines 55--60 to clarify both features
 of the method.

@@

In the field of connectivity analysis, the issue of distortion has posed important problems in the
  standard methodologies to build the connectivity network\footnote{generally imposing
  a parcellation mapped from T1-space}.
With the emergence of novel reconstruction and tractography methods (see lines 5--15),
  these tasks may also be improved with the application of more accurate registration
  methods like \emph{regseg}.

The given references are relevant to the domain of diffusion image processing,
  specially that provided as [1], and propose the following discussion that
  is very interesting but goes beyond the objectives of this paper.
\cite{jones_twentyfive_2010} present a list of common pitfalls in the
  analysis of dMRI.
Particularly, pitfalls \#1, \#2, and \#14 apply to our work:
\begin{itemize}
\item Pitfall \#1 points to the problem of modulation in the unwrapped signal and the
  appearance of unrealistic structures due to interpolation.
Since we promote to perform the analysis in the original space, this problem does not 
  apply in practice.
\item Closely related, pitfall \#2 raises the problem of misplaced tracks (and cites
  the other two references given [2, 3]).
We have included in the discussion (see lines 343-344) some text indicating that the tracks
  can be unwarped using the deformation found with \emph{regseg} since it is revertible.
However, checking the correctness of tracks computed on unwarped dMRI data vs. tracks computed
  on the original dMRI data and then unwarped with \emph{regseg} or whatever other
  technique is out of the scope of the paper, but certainly a very interesting lead for
  future research.
As a matter of fact, we have recently published a whole-brain digital phantom of dMRI 
  data \citep{esteban_diffantom_2016} to perform evaluations like the one suggested 
  here \citep[for instance, please see][Chapter 5]{esteban_image_2015}.
\item Finally, pitfall \#14 has side relationships with our method, since \emph{regseg} is
  definitely intended to increase the accuracy of the ROIs imposed on the dMRI data.
In this case, Jones et al. blame it on the misregistration of the \emph{b0} and the
  diffusion weighted volumes, but the problem applies equally when the ROIs are
  not correctly aligned with the \emph{b0} itself.

Regarding how the method integrates within the diffusion processing pipeline,
  we have included the new lines 339--344.
\end{itemize}
}

@@

\begin{reviewpoint}
Pg 5 regularisation: `` requires that the distortion and its gradient have zero mean and variance governed by the matrices A and B'' suggest adding a comma here for easier interpretation i.e. ``requires that the distortion and its gradient have zero mean, and variance governed by the matrices A and B''
\end{reviewpoint}
\reply{%
We appreciate the suggestion and introduced the comma accordingly.
}

\begin{reviewpoint}
Pg 7, line 4 `` anisotropy is aligned''. What is meant by anisotropy in this context, the anisotropy of the deformation gradient.
\end{reviewpoint}
\reply{%
The reviewer is totally right. We have modified the text accordingly (lines 116--117,
  and lines 400--402 in the Appendix 1). 
This comment is close and consistent with \autoref{com:31}.
}

@@

\begin{reviewpoint}
Figure 2. This looks nice, but its not easy to interpret. I can't see a clear improvement through AC. I think there needs to be direct comparisons between the conditions represented by AC on one plot. Why not have 6 plots for each tissue class comparing distributions for each condition.
\end{reviewpoint}
\reply{%
We have modified the figure and its caption as suggested.
}
% ------------------------------------------------------------------- Reviewer 3
\bigskip
\bigskip
\subsection*{Reviewer \#3:}
\begin{reviewpoint}
This paper presents a new method for correction of EPI distortion in diffusion MRI data by performing simultaneous registration and segmentation of a set of nested surfaces, using a multichannel FA/MD image with a registration target of an undistorted T1 image. The primary advantage over other methods is that cortical surfaces and whitegray matter boundaries are apparently aligned well by this method. Results show that it outperforms one implementation of the more standard baseline to T2 registration. Overall this is a quite interesting work, and it will be a good addition to the literature and also to the body of opensource code for diffusion MRI.

I have some comments for clarification of the paper.
\end{reviewpoint}
\reply{%
We thank the reviewer for the time dedicated in his/her review and these positive comments.
Again, we are pleased to confirm that the summary of the reviewer agrees with our objectives in this
  work.
}

\begin{reviewpoint}
Overall, I am slightly confused by some aspects of the equations, and I think that some information from the supplement (which I have not had time to read) should be moved into the paper to clarify missing points. The paper should be able to be read independently of the supplement, which does not seem to be the case currently. Specifically, the relationship of the bspline registration framework to the evolution of the active contours is a bit confusing, with a main equation given inside Figure 1, and apparently all other detail in the supplement. This could be as simple as an additional paragraph or two.
\end{reviewpoint}
\reply{%
We have introduced one Appendix (\emph{Appendix 2. Application of the shape-gradients}, starting by line 402) to
  show how the shape-gradients are applied and how the equation given
  in the Figure 1 relates to the registration framework.
As the reviewer indicated, this text was part of the Supplemental Materials.
}

@@

\begin{reviewpoint}
Please state that the goal of the paper is to improve connectomebased diffusion MRI research. The processing description on pages 12 assumes that the connectomebased approach is the only approach to analysis of diffusion MRI. But this is clearly not the case (surgical planning, TBSS, etc).
\end{reviewpoint}
\reply{%
The reviewer is right in the sense that our method was devised for its application on the extraction of
  the human connectome from dMRI data.
However, and with certain adaptations, the registration method is also useful in other approaches such as those
  suggested.
Therefore, we have relaxed those statements restricting the use of this method to the connectome
  extraction (in the abstract, and lines 323--324) and included brief indications on how this could be applied in
  other dMRI applications (lines 343--344).
}

\begin{reviewpoint}
``to current models such as (Dadicci et al)'' is confusing. Please give the name(s) of the current model(s).
\end{reviewpoint}
\reply{%
We have introduced the name of the model referred to (line 8).
}

\begin{reviewpoint}
On page 3, please define N (section 2.1)
\end{reviewpoint}
\reply{%
We have introduced the symbol $\mathcal{N}$ before eq. (4) in which it appears for first time (line 89).
}

\begin{reviewpoint}
In Figure 1, it is a bit confusing that active contours appear, because the abstract did not mention active contours directly, and the evolution following the inward normals at each vertex is not explicit in the equations of the paper.
\end{reviewpoint}
\reply{%
The active contours are explicitly introduced in line 46, and an implicit reference to this methodological foundation
is found in the abstract: ``\emph{we propose regseg, a surface-to-volume nonlinear registration method that segments homogeneous regions within multivariate images by mapping a set of nested reference-surfaces}''.
The reviewer is right regarding the missing details about the direction of normals.
In the revised manuscript, this is described in the new Appendix 1 (equations A.1--4), and the caption of
  Figure 1 has been also modified.
}

\begin{reviewpoint}
It seems the evolution at $s_1$ is in the direction opposite the inward normal, so the caption of Figure 1 is a bit confusing, since it directly states the inward normals are followed, rather than just the normal to the curve.
\end{reviewpoint}
\reply{%
The evolution at $s_1$ can be in either direction, depending on the direction of the gradient.
Again, with modification of the caption of Fig. 1 and the insertion of Appendix 1 we understand that this issue is clarified.
}

@@

\begin{reviewpoint}
I recommend adding a little bit of text to unify figure 1 and the surface evolution framework with the concepts presented in the registration equations section. I think this information may be in the supplement.
\end{reviewpoint}
\reply{%
We have included two equations (A.6, A.7) from the Supplemental Materials in the new Appendix 1.
Additionally, some text has been added (lines 122--125) to ensure the necessary consistency and connect the formulations in
  the manuscript to the Figure 1.
}

\begin{reviewpoint}
On page 5, equation (4), I think f should be f prime on the first line.
\end{reviewpoint}
\reply{%
The reviewer is right. We have fixed the manuscript accordingly.
}

\begin{reviewpoint}
Please describe A and B in a sentence after equation 5. It is not clear how these were chosen (or at least refer to the appendix).
\end{reviewpoint}
\reply{%
Added a sentence describing A and B (lines 102--103).
}

@@

Regarding the last part of the question, the second integral is defined directly along the whole image
  domain $\Omega$ because it does not depend on the corresponding partition $\{\Omega_l\}$.
This integral corresponds to the regularization term.
The first integral is computed region-wise, since the energy functional has different
  parameters $\{\boldsymbol{\mu}_l, \boldsymbol{\Sigma}_l\}$ for each of the regions $\Omega_l$.
}

\begin{reviewpoint}
In equation (7) please clarify with brackets what the sum over l is applied to.
\end{reviewpoint}
\reply{%
Added large braces.
}

\begin{reviewpoint}
The two pluses are present again in equation (10).
\end{reviewpoint}
\reply{%
Added brackets as in \autoref{com:19} and fixed two pluses as in \autoref{com:18}.
}

\begin{reviewpoint}
Please explain in a sentence what alpha and beta are in equation (10) (expected variances along each axis of the coordinate system, or something).
\end{reviewpoint}
\reply{%
Added explanation in lines 119--120.
}

\begin{reviewpoint}
$\mathbf{g}_k$ is completely undefined in equation (11). I see there is information in a supplement but at a minimum each variable should be defined in a sentence in the main paper.
\end{reviewpoint}
\reply{%
Equations (A.6) and (A.7) introduced in the new Appendix 2 define $\mathbf{g}_k$ (starts by line 423).
Additionally, a reference to the Appendix and corresponding equations (lines 123--124) and a brief
  explanation of $\mathbf{g}_k$ (line 125) have been added.
}

@@

\begin{reviewpoint}
Figure 6. Are the plots in B from one or many subjects? Please state in the caption.
\end{reviewpoint}
\reply{%
We have updated the caption accordingly.
}

\begin{reviewpoint}
In Table 2 caption how many subjects were input to this experiment? This is given as 16 in the abstract, but that info is not very near the actual results and would increase their impact.
\end{reviewpoint}
\reply{%
We have updated the caption accordingly.
}

\begin{reviewpoint}
pg 13. ``We also show the 95\% CIs of the sWI for these surfaces.'' sounds like it is referring to Figure 6. However it is apparently in Table 2, which should be referenced in this sentence. Or the sentences about Table 2 should be reorganized to sound separate from Figure 6.
\end{reviewpoint}
\reply{%
The reference to Table 2 has been added to that sentence (line 249).
}

\begin{reviewpoint}
in Appendix (and also elsewhere): ``the anisotropy is aligned with the imaging axes'' is quite confusing. Usually anisotropy refers to FA or similar. I believe this is the anisotropy of the covariance matrices, really that their eigenvectors are assumed aligned with the voxel coordinate system. Please clarify the usage of ``anisotropy'' in the registration framework.
\end{reviewpoint}
\reply{%
The reviewer is totally right. We have modified the text accordingly (lines 116--117,
  and lines 400--402 in the Appendix 1).
This comment is also made by the reviewer \#1 in the \autoref{com:5}.
}

\begin{reviewpoint}
What values are used for alpha and beta in the appendix?
\end{reviewpoint}
\reply{%
We agree with the reviewer on the interest of the actual values for $\boldsymbol{\alpha}$
  and $\boldsymbol{\beta}$ to replicate our experiments or derive new works.
We have included the typical range for these parameters in lines 131--133.
In order to ensure the reproducibility of the presented experiments, the values
  used in the experiments are provided in the open source repository in the 
  human-readable JSON file format.
Please find the references to the parameters files in the footnotes of pages 9 and 10.
}

@@

pg 2: ``threshold set'' sounds like a mathematical set > ``threshold that is chosen''

pg 5. ``displacements field'' > ``displacement field''

pg. 15 ``monitorization'' is not a word > monitoring

\end{reviewpoint}
\reply{%
All the issues have been corrected following the suggestions of the reviewer.
In particular, som additional appearances of ``displacements field'' have been corrected.
}

### 2016-02-03 Publications/2015-NeuroImage/03-introduction.tex
revised paper

Here, we present a segmentation and surface-to-volume registration method
  named \regseg{}, and show its usefulness in mapping anatomical information from structural
  space into native \gls*{dmri} space to aid subsequent processing steps
  (reconstruction, tractography and network building using a cortical parcellation).
The underlying hypothesis is that the registration and segmentation problems
  in \gls*{dmri} can be solved simultaneously.
To implement \regseg{} we first establish an active-contours without edges
  \citep{chan_active_2001} segmentation framework.
A specific set of reference surfaces extracted from the same subject initialize
  the 3D active contours, which evolve searching for homogeneous regions in the multivariate
  target-image.
In the presented application to segment \gls*{dmri} data, we propose to use \regseg{}
  using a set of nested surfaces extracted from a structural image (e.g. \gls*{t1}) and
  a bivariate target-volume comprehending the \gls*{fa} and \gls*{adc} maps as inputs.
The evolution of the surfaces is supported by a B-spline basis, optimized
  iteratively using a descent approach driven by shape-gradients
  \citep{besson_dream2s_2003,herbulot_segmentation_2006}.
Therefore, \regseg{} establishes a registration framework that actually
  deals with the nonlinear warping induced by \gls*{epi} distortions.
\Regseg{} integrates the benefits of segmentation and registration methods together and
  exploits the multivariate nature of \gls*{dmri} data to contribute in the proposed
  application on three key aspects:
  1) the surfaces are typically extracted from the \gls*{t1} image of the same subject,
    therefore \regseg{} does not require additional MR acquisitions to the minimal
    \gls*{dmri} protocol \revcomment[R\#1-C.2]{in order to estimate the deformation field};
  2) \revcomment[R\#1-C.2]{depending on the application, the resampling of the \glspl*{dwi}
    introduced by the correction method can be avoided, either by performing the posterior
    processing on the native \gls*{dmri} space or applying the unwarping on the tractography
    itself}; and
  3) \regseg{} increases the geometrical accuracy of the overall process.
In this paper, we first verify the functionality of the method and the \regseg{}
  implementation using a set of digital phantoms, demonstrating the sub-pixel accuracy
  in registration.
We then evaluate \regseg{} on real \gls*{dmri} datasets, using a derivation of our
  instrumentation framework \citep{esteban_simulationbased_2014} which simulates
  known and realistic \gls*{epi} distortions.
We also compare \regseg{} and a nonlinear registration method to map the \lowb{} to
  the corresponding \gls*{t2} image of the same subject.
This method is a derivation of the \gls*{t2b} correction methods \citep{kybic_unwarping_2000},
  reproducing the settings of an existing implementation included within a widely used diffusion
  processing software \citep[\emph{ExploreDTI},][]{leemans_exploredti_2009}.
With this comparison, we demonstrate how \regseg{} achieves higher accuracy with the
  simultaneous registration and segmentation process.

>>>>

Here, we present a segmentation and surface-to-volume registration method
  called \regseg{}, and show its usefulness in mapping anatomical information from structural
  space into native \gls*{dmri} space to aid subsequent processing steps
  (reconstruction, tractography and network building using a cortical parcellation).
The underlying hypothesis is that the registration and segmentation problems
  in \gls*{dmri} can be solved simultaneously.
To implement \regseg{} we first establish an active-contours without edges
  \citep{chan_active_2001} segmentation framework.
A specific set of reference surfaces extracted from the same subject initialize
  the 3D active contours, which evolve searching for homogeneous regions in the multivariate
  target-image.
\revcomment[authors]{We apply \regseg{} to segment \gls*{dmri} data by mapping
  a} set of nested surfaces extracted from a structural image (e.g. \gls*{t1}) to
  a bivariate target-volume comprehending the \gls*{fa} and \gls*{adc} maps.
The evolution of the surfaces is supported by a B-spline basis, optimized
  iteratively using a descent approach driven by shape-gradients
  \citep{besson_dream2s_2003,herbulot_segmentation_2006}.
Therefore, \regseg{} establishes a registration framework that actually
  deals with the nonlinear warping induced by \gls*{epi} distortions.
\Regseg{} integrates the benefits of segmentation and registration methods together and
  exploits the multivariate nature of \gls*{dmri} data to contribute in the proposed
  application on three key aspects:
  1) the surfaces are typically extracted from the \gls*{t1} image of the same subject,
    therefore \regseg{} does not require additional MR acquisitions to the minimal
    \gls*{dmri} protocol \revcomment[R\#1-C.2]{in order to estimate the deformation field};
  2) \revcomment[R\#1-C.2]{depending on the application, the resampling of the \glspl*{dwi}
    introduced by the correction method can be avoided, either by performing the posterior
    processing on the native \gls*{dmri} space or applying the unwarping on the tractography
    itself}; and
  3) \regseg{} increases the geometrical accuracy of the overall process.
In this paper, we first verify the functionality of the method and the \regseg{}
  implementation using a set of digital phantoms, demonstrating the sub-pixel accuracy
  in registration.
Then, we evaluate \regseg{} on real \gls*{dmri} datasets, using a derivation of our
  instrumentation framework \citep{esteban_simulationbased_2014} which simulates
  known and realistic \gls*{epi} distortions.
We also compare \regseg{} and a nonlinear registration method to map the \lowb{} to
  the corresponding \gls*{t2} image of the same subject.
\revcomment[authors]{This approach is the first step of the} \gls*{t2b} correction
  methods \citep{kybic_unwarping_2000}.
\revcomment[authors]{We reproduce the settings and implementation of a} widely used diffusion
  processing software \citep[\emph{ExploreDTI},][]{leemans_exploredti_2009}.
With this comparison, we demonstrate how \regseg{} achieves higher accuracy with the
  simultaneous registration and segmentation process.
